Retrofitting Effect Handlers onto OCaml

### 2020-11-20 retro-concurrency.tex
addition of branch predictor and stack overflow check sentence to heap-allocated fibers section
Some rewording to free up some space and make things tighter

We have implemented effect handlers in a multicore extension of the OCaml
programming language. We call our implementation \textbf{Multicore OCaml} to
distinguish it from stock OCaml.

The remainder of the paper continues with a description of the stock OCaml
program stack (\S\ref{sec:stack}). We then refine the design further by
focussing on retrofitting effect handlers into a mainstream \emph{systems}
language (\S\ref{sec:refine}), followed by the compilation scheme and
operational semantics for Multicore OCaml effect handlers
(\S\ref{sec:semantics}).  We then discuss the compiler and runtime system
support for implementing effect handlers (\S\ref{sec:impl}), and present
extensive performance evaluation of effect handlers (\S\ref{sec:eval}) against
our design goals (\S\ref{sec:req}). Finally, we discuss the related work
(\S\ref{sec:related}) and conclude (\S\ref{sec:conc}).

@@

Next, we have the variable-sized area for the OCaml frames. In order to keep
fibers small, this area is initially 16 words in length. When the stack pointer
|rsp| becomes less than the \emph{stack threshold} (maintained in the
|Caml_state| table), the stack is said to have overflown. On stack overflow, we
copy the whole fiber to a new area with double the size. In Multicore OCaml, we
introduce stack overflow checks into the function prologue of OCaml functions.

@@

\textbf{\textit{Generators.}} Generators allow data structures to be traversed on
demand. Many languages including JavaScript and Python provide generators as a
built-in primitive. Using effect handlers (|MC|), given any data structure
(|'a t|) and its iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can
derive its generator function (|val next : unit -> 'a option|). The code is
available in the supplementary material. We evaluate the performance of
traversing a complete binary tree of depth 25 using this generator. This
involves $2^{26}$ stack switches in total. For comparison, we also implemented
a hand-written, selective CPSed~\cite{Nielson01},
defunctionalised~\cite{Danvy01} version (|cps|) and a concurrency monad
(|monad|) version of the generator for the tree. Both |cps| and |monad|
versions are specialised to the binary tree, and come with the usual caveats of
not using the stack for function calls. We observed that the |cps| version was
the fastest, thanks to specialisation and hand optimisation. |MC| version was
only 2.76$\times$ slower than |cps| while being a generic
solution, whereas the |monad| version was 8.69$\times$ slower than |cps|.

@@

Using effect handlers in the web server also helps with debugging the system as
a whole. Multicore OCaml supports getting the backtrace of a continuation,
which is a minor extension to getting the current backtrace. We can use this
feature to dump the backtrace of all the user-level threads in the web server.
Such a feature is provided by Go~\cite{gopprof}, but not possible with OCaml
concurrency libraries such as Async and Lwt which lack the notion of a thread.

@@

There are several strategies for implementing effect handlers. Eff~\cite{Eff},
Helium~\cite{Biernacki20}, Frank~\cite{Frank} and the Links server
backend~\cite{Hillerstrom20} use an interpreter similar to our operational
semantics to implement effect handlers. Effekt~\cite{Effekt}, Links JavaScript
backend~\cite{Hillerstrom20} and Koka~\cite{Leijen17} use type-directed
selective CPS translation. These language are equipped with an effect system,
which allows compiling pure code in direct-style and effectful code in CPS.
Leijen~\cite{Leijen14} implements effect handlers as a library in C using stack
copying. Since C allows pointers into the stack, care is taken to ensure that
when continuations are resumed, the constituent frames are restored to the same
memory addresses as at the time of capture. Kiselyov et al.~\cite{Kiselyov18}
use an implementation of multi-prompt delimited continuations as an OCaml
library~\cite{Kiselyov12} to embed the Eff language in OCaml. Indeed, Forester et
al.~\cite{Forster19} showed that in an untyped setting, effect handlers,
monadic reflection and delimited control can macro-express each other.

Multicore OCaml uses the call stack for implementing continuations (as
do~\cite{Leijen14, Kiselyov12}), but with one-shot continuations. Bruggeman et
al.~\cite{Bruggeman96} show how to implement one-shot continuations efficiently
using segmented stacks in Scheme. Farvardin et al.~\cite{Farvardin20} perform a
comprehensive evaluation of various implementation strategies for continuations
on modern hardware. Multicore OCaml stacks do not neatly fit the description of
one of these implementation strategies -- they are best described as using the
resize strategy from Farvardin et al. for each of the fibers, which are in turn
linked to represent the current stack and the captured continuations.

@@

Multicore OCaml builds on the existing defensive coding practices against
exceptions to clean up resources on non-returning exits. Our continuations are
assumed to be resumed exactly once using |continue| or |discontinue|. As a
result, when a computation performs an effect, we are guaranteed that the
control will return. For the non-returning cases (value and exceptional
return), the code already handles resource cleanup. OCaml does not have a
\texttt{\footnotesize try/finally} construct commonly used for resource cleanup
in many programming languages. Alternative standard libraries such as
Base~\cite{BaseProtect} and Core~\cite{CoreProtect} provide mechanisms
analogous to |unwind-protect|, which are in turn implemented using exceptions.
Thus, the linear use of continuations enabled by the |discontinue| primitive
ensures backwards compatibility of legacy OCaml systems code under non-local
control flow introduced by effect handlers. Leijen~\cite{Leijen18} explicitly
extends effect handlers with |initially| and |finally| clauses in Koka for
resource safety. Dolan et al.~\cite{TFP17} describe the interaction
of effect handlers and asynchronous exceptions. This is orthogonal to the
contributions of this paper. Our focus is the compiler and runtime system
support for implementing effect handlers.

>>>>

We have implemented effect handlers in a multicore extension of the OCaml
programming language which we call \textbf{Multicore OCaml} to distinguish
it from stock OCaml.

The remainder of the paper continues with a description of the stock OCaml
program stack (\S\ref{sec:stack}). We then refine the design further by
focussing on retrofitting effect handlers into a mainstream \emph{systems}
language (\S\ref{sec:refine}), followed by the compilation scheme and
operational semantics for Multicore OCaml effect handlers
(\S\ref{sec:semantics}).  We then discuss the compiler and runtime system
support for implementing effect handlers (\S\ref{sec:impl}), and present an
extensive performance evaluation of effect handlers (\S\ref{sec:eval}) against
our design goals (\S\ref{sec:req}). Finally, we discuss the related work
(\S\ref{sec:related}) and conclude (\S\ref{sec:conc}).

@@

Next, we have the variable-sized area for the OCaml frames. In order to keep
fibers small, this area is initially 16 words in length. When the stack pointer
|rsp| becomes less than the \emph{stack threshold} (maintained in the
|Caml_state| table), the stack is said to have overflown. On stack overflow, we
copy the whole fiber to a new area with double the size. In Multicore OCaml, we
introduce stack overflow checks into the function prologue of OCaml functions.
These stack overflow operations are rare and so will be correctly predicted by
the CPU branch predictor.

@@

\textbf{\textit{Generators.}} Generators allow data structures to be traversed on
demand. Many languages including JavaScript and Python provide generators as a
built-in primitive. Using effect handlers (|MC|), given any data structure
(|'a t|) and its iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can
derive its generator function (|val next : unit -> 'a option|). The code is
available in the supplementary material. We evaluate the performance of
traversing a complete binary tree of depth 25 using this generator. This
involves $2^{26}$ stack switches in total. For comparison, we also implemented
a hand-written, selective CPSed~\cite{Nielson01},
defunctionalised~\cite{Danvy01} version (|cps|) and a concurrency monad
(|monad|) version of the generator for the tree. Both |cps| and |monad|
versions are specialised to the binary tree with the usual caveats of
not using the stack for function calls. We observed that the |cps| version was
the fastest, thanks to specialisation and hand optimisation. |MC| version was
only 2.76$\times$ slower than |cps| while being a generic
solution, whereas the |monad| version was 8.69$\times$ slower than |cps|.

@@

Multicore OCaml supports backtraces for continuations in addition to backtraces
of the current stack as in stock OCaml. Using effect handlers in a system such
as a web server aides debugging and profiling because it is possible to get a
backtrace snapshot of all current requests. Such a feature is available in
Go~\cite{gopprof}, but not in OCaml concurrency libraries such as Async and
Lwt which lack the notion of a thread.

@@

There are several strategies for implementing effect handlers. Eff~\cite{Eff},
Helium~\cite{Biernacki20}, Frank~\cite{Frank} and the Links server
backend~\cite{Hillerstrom20} use an interpreter similar to our operational
semantics to implement effect handlers. Effekt~\cite{Effekt}, Links JavaScript
backend~\cite{Hillerstrom20} and Koka~\cite{Leijen17} use type-directed
selective CPS translation. These language are equipped with an effect system,
which allows compiling pure code in direct-style and effectful code in CPS.
Leijen~\cite{Leijen14} implements effect handlers as a library in C using stack
copying. C allows pointers into the stack, so care is taken to ensure that
when continuations are resumed, the constituent frames are restored to the same
memory addresses as at the time of capture. Kiselyov et al.~\cite{Kiselyov18}
use an implementation of multi-prompt delimited continuations as an OCaml
library~\cite{Kiselyov12} to embed the Eff language in OCaml. Indeed, Forester et
al.~\cite{Forster19} showed that in an untyped setting, effect handlers,
monadic reflection and delimited control can macro-express each other.

Multicore OCaml uses the call stack for implementing continuations (as
do~\cite{Leijen14, Kiselyov12}), but with one-shot continuations. Bruggeman et
al.~\cite{Bruggeman96} show how to implement one-shot continuations efficiently
using segmented stacks in Scheme. Farvardin et al.~\cite{Farvardin20} perform a
comprehensive evaluation of various implementation strategies for continuations
on modern hardware. Multicore OCaml stacks do not neatly fit the description of
one of these implementation strategies -- they are best described as using the
resize strategy from Farvardin et al. for each of the fibers, which are
linked to represent the current stack and the captured continuations.

@@

Multicore OCaml builds on the existing defensive coding practices against
exceptions to clean up resources on non-returning exits. Our continuations are
assumed to be resumed exactly once using |continue| or |discontinue|. As a
result, when a computation performs an effect, we are guaranteed that the
control will return. For the non-returning cases (value and exceptional
return), the code already handles resource cleanup. OCaml does not have a
\texttt{\footnotesize try/finally} construct commonly used for resource cleanup
in many programming languages. Alternative standard libraries such as
Base~\cite{BaseProtect} and Core~\cite{CoreProtect} provide mechanisms
analogous to |unwind-protect|, which are implemented using exceptions.
Thus, the linear use of continuations enabled by the |discontinue| primitive
ensures backwards compatibility of legacy OCaml systems code under non-local
control flow introduced by effect handlers. Leijen~\cite{Leijen18} explicitly
extends effect handlers with |initially| and |finally| clauses in Koka for
resource safety. Dolan et al.~\cite{TFP17} describe the interaction
of effect handlers and asynchronous exceptions. This is orthogonal to the
contributions of this paper. Our focus is the compiler and runtime system
support for implementing effect handlers.
